---
title: "OutTakes.Rmd"
author: "Jim Callahan"
date: "October 20, 2015"
output: html_document
---
OutTakes: stuff prepared for, but did not make it into final report.

This study does not show transmission type 
(automatic vs. manual) to be significant once one accounts for weight and 
number of engine cylinders. In fact, mpg may be accurately modeled simply using a
transformed version of weight to allow for non-linear effects.

This result, however, may represent a flaw in the study design; this regression study, 
relies on group averages and not paired data. Paired data would be closer 
to the consumer experience of evaluating one model of car with (standard) manual 
or (optional) automatic transmission. 
In other words, there may be a more distinct effect when one examines one model of car at a time
with manual or automatic transmission rather than pooling several models of cars 
together in one data set. Unfortunately, the mtcars dataset does not contain identical
models of cars with and without an optional automatic transmission. So, a paired
test could not be conducted with this data. So, for example, this study is accounting
for the difference in mpg between a Toyota Corona and a Cadillac Fleetwood and
**not** the difference between a Toyota Corona with standard (manual) transmission versus
a Totyota Corna with the automatic transmission option.

```{r}
# Separate regressions of mpg on weight for each number of cylinders
# create factors with value labels 
library(ggplot2)
data(mtcars)
mtcars$gear <- factor(mtcars$gear,levels=c(3,4,5),
                      labels=c("3gears","4gears","5gears")) 
mtcars$am <- factor(mtcars$am,levels=c(0,1),
                    labels=c("Automatic","Manual")) 
mtcars$cyl <- factor(mtcars$cyl,levels=c(4,6,8),
                     labels=c("4 cylinder","6 cylinder","8 cylinder")) 

# All -- use everything
MPGModAll <- lm(mpg  ~ . , data = mtcars); 
summary(MPGModAll)

# Weight is significant
MPGmod001 <- lm(mpg ~ as.factor(am)+wt, data=mtcars)
summary(MPGmod001)

# Cylinders helps
MPGmod002 <- lm(mpg ~ as.factor(am)+wt+as.factor(cyl), data=mtcars)
summary(MPGmod002)

# drop Transmission (am) *** BEST MODEL ***
par(mfrow = c(1,2))
MPGmod003 <- lm(mpg ~ wt+as.factor(cyl), data=mtcars)
summary(MPGmod003)
plot(MPGmod003, which = 1)
plot(MPGmod003, which = 2)

MPGmod005 <- lm(mpg ~ log(wt), data=mtcars)
summary(MPGmod005)

MPGmod005 <- lm(mpg ~ log(wt)+as.factor(cyl), data=mtcars)
summary(MPGmod005)

qplot(wt, mpg, data=mtcars, geom=c("point", "smooth"), 
      method="lm", formula=y~x, color=cyl, 
      main="Regression of MPG on Weight by Engine Cylinders
    from QuickR blog post by Robert I. Kabacoff", 
      xlab="Weight (1,000 lbs)", 
      ylab="Miles per Gallon - MPG")

qplot(wt, mpg, data=mtcars, geom=c("point", "smooth"), 
      method="lm", formula=y~log(x), color=cyl, 
      main="Regression of MPG on Weight by Engine Cylinders
    from QuickR blog post by Robert I. Kabacoff", 
      xlab="Weight (1,000 lbs)", 
      ylab="Miles per Gallon - MPG")
```

```{r}
qplot(wt, mpg, data=mtcars, geom=c("point", "smooth"), 
      method="lm", formula=y~I(x^2)+x, color=cyl, 
      main="Regression of MPG on Weight by Engine Cylinders
    from QuickR blog post by Robert I. Kabacoff", 
      xlab="Weight (1,000 lbs)", 
      ylab="Miles per Gallon - MPG")
```

```{r}
qplot(wt, mpg, data=mtcars, geom=c("point", "smooth"), 
      method="lm", formula=y~I(x^2)+x, 
      main="Regression of MPG on Weight by Engine Cylinders
    from QuickR blog post by Robert I. Kabacoff", 
      xlab="Weight (1,000 lbs)", 
      ylab="Miles per Gallon - MPG")
```

```{r}
qplot(wt, mpg, data=mtcars, geom=c("point", "smooth"), 
      method="lm", formula=y~log(x), 
      main="Regression of MPG on Weight by Engine Cylinders
    from QuickR blog post by Robert I. Kabacoff", 
      xlab="Weight (1,000 lbs)", 
      ylab="Miles per Gallon - MPG")
```

```{r}
attach(mtcars)
# wt = Weight (lb/1000)
summary( wt )
# cyl = Number of Cylinders
# Count of cars by Number of Cylinders
summary( as.factor(cyl) )
# mpg by Number of Cylinders
tapply(mpg, as.factor(cyl), mean )

# disp = Displacement (cu.in.)
summary(disp )
detach(mtcars)
```
